---
title: "HW 7 - Numerical Integration"
author: "Sophie Shan"
date: "03/15/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
#set working directory 
setwd("/Users/sophieshan/Desktop/OneDrive - University of North Carolina at Chapel Hill/UNC/Spring 2024/BIOS 735/Homework Assignments/HW7/")

#load relevant libraries 
#install.packages("statmod")
library(statmod)
```

# Maximization of poisson GLMM from lecture

Now that we have discussed several approaches for numerical integration, lets now maximize the model given in lecture.  You may choose any maximization approach, as well as any numerical integration procedure from lecture, to obtain the MLE's for $\boldsymbol{\beta}$ and $\sigma_{\gamma}^2$.  

Hint: You should evaluate a number of intervals/nodes, decreasing the convergence threshold, etc and evaluate its impact before reporting your final result. We have shown how to perform, for example, AGQ and IS to obtain the likelihood pertaining to the first subject from class. 

```{r, warnings = FALSE}
## Solution: place relevant helper functions pertaining to integration here 

#From the lecture 
inner = function(gammai, yi, xi = cbind(rep(1,5), 1:5), beta, s2gamma, log = T) {

  ## create vector holding return value f(gammai) with respect to each element of gammai
    val = rep(NA, length(gammai))
  
  ## calculate product with respect to each element of gammai
    for (i in 1:length(val)) {
      
        # calculate lambda:  x_{ij} %*% beta + gammai[i]
        # 5 x 1 vector pertaining to each month's measurement
        lambda = exp(xi %*% beta + gammai[i])
        
        # calculate f(y_ij | lambda_ij) for j = 1,...,5
        # 5 x 1 vector
        val0 = dpois(x = yi, lambda = lambda)
        
        # calculate the product of f(y_ij | lambda_ij) over
        # j = 1,...,5 times phi(gamma_i | s2_gamma)
        # \prod_{j = 1}^{5}f(y_{1j} | \lambda_{1j})\phi(\gamma_1| 0, \sigma^2_{\gamma})
        val[i] = prod(val0) * dnorm(x = gammai[i],
                                    mean = 0,
                                    sd = sqrt(exp(s2gamma)))
      }
  
  ## return likelihood or log likelihood pertaniing to i'th subject
    if (log == F) {
      return(val)
    } else{
      return(log(val))
    }
    
}

trap = function(intervals, range, yi, beta, s2gamma, trace = 0){

  ## set the number of intervals n
    n = intervals
    
  ## evaluate Reimann’s rule, assuming intervals of equal length
    # get h
    h = (range[2] - range[1])/n
  
    # now evaluate.  The book says to sum from 0 to n-1.  
    # 0 below pertains to the left most interval boundary
    k = 0:(n-1)
  
    # we need to adjust this for how the r vector 
    # indices are specified (add 1 since no 0 index in r)
    # now two nodes per index, left and right boundary
    points_left = range[1] + k * h
    points_right = range[1] + (k + 1) * h
    
    # now evaluate the trap rule for equal intervals as described above
    # f(left) first
    left  = inner(
      gammai = points_left,
      yi = yi,
      beta = beta,
      s2gamma = s2gamma
    )
    
    # then f(right)
    right  = inner(
      gammai = points_right,
      yi = yi,
      beta = beta,
      s2gamma = s2gamma
    )
    
    # calculate value, following book formula
    val = (h / 2) * sum(left + right)
  
  ## return the result
    return(val)
}

## End Solution



## Solution: place relevant helper functions pertaining to maximization here (likelihood etc)

logLik <- function(x, y, intervals, range){
  
  beta = x[1:2]
  s2gamma = x[3]
  ll = 0
  for(i in 1:max(y$subject)){
    ll = ll + trap(intervals = intervals, range = range, yi = y$words[y$subject == i], beta = beta, s2gamma = s2gamma, trace = 1)
  }
  
  return(ll)
  
}


## End Solution



## Solution: place primary code for maximization here, calling functions in the above two sections
## Remember to print your primary results and use the following starting values
beta = c(1.804, 0.165)
s2gamma = log(0.000225) #use this because sometimes taking sqrt(s2gamma) fails b/c its too small
alz = read.table("alzheimers.dat", header = T)

fit = optim(
  par = c(beta, s2gamma),
  fn = logLik,
  method = "Nelder-Mead",
  y = alz,
  intervals = 1000,
  range = c(-0.07, 0.085),
  control = list(fnscale = -1) #default is minimization
)



## End Solution

#betas:
fit$par[1:2]

#sigma
exp(fit$par[3])

```

# Plot

Now, plot the fitted line from the fitted GLMM on the spaghetti plot from lecture

```{r}
## solution

#####################
#Spaghetti Plot From Lecture 
#####################
## plot trajectory of each patient as separate lines
  for (i in 1:max(alz$subject)) {
    # select subject indices
    index = which(alz$subject == i)
    
    # plot trajectory
    if (i == 1) {
      plot(
        alz$month[index],
        alz$words[index],
        type = 'l',
        ylim = range(alz$words),
        ylab = "Words",
        xlab = "Month",
        col = i
      )
    } else{
      lines(alz$month[index], alz$words[index], type = 'l', col = i)
    }
  }

lines(1:5, exp(fit$par[1] + fit$par[2]*1:5), lwd = 5)
## end solution
```
